---
title: "Data_Formatting_PreOp"
author: "Celeste Basken"
date: "2026-04-08"
output: html_document
---

The purpose of this doc is just to take the Ingredient_prices, GHG_equivalents, and F25_Sp26_meals, and then build the cleaned version of the df which is optimization-ready. 

This used to be done in google sheets (for example, https://docs.google.com/spreadsheets/d/14wqCazTvzhN5UJX_mCnG3Jjdsi1-8abxkL5vip48bFU/edit?gid=763421786#gid=763421786), but that's actually super inefficient and it sucks, so now I'm doing it all in r. 

To change out the data, simply edit the base spreadsheets.

Packages:
```{r}
library(dplyr)
library(readr)
library(janitor)
```

## Read Data

```{r}
safe_mean <- function(x) {
  if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
}

first_nonmissing <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) NA_character_ else x[1]
}

# -------------------------
# Read + clean
# -------------------------
meals <- read.csv("../Basic_Data/F25_Sp26_meals.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ingredient_prices <- read.csv("../Basic_Data/Ingredient_prices.csv", stringsAsFactors = FALSE) %>%
  clean_names()

ghg_equivalents <- read.csv("../Basic_Data/GHG_equivalents.csv", stringsAsFactors = FALSE) %>%
  clean_names()

# -------------------------
# Standardize values
# -------------------------
meals <- meals %>%
  mutate(
    dining_hall = trimws(dining_hall),
    ingredient = trimws(ingredient),
    category = trimws(category),
    planned_portions = as.numeric(gsub(",", "", planned_portions)),
    percent_meat = as.numeric(percent_meat),
    meat_price_per_dish = as.numeric(meat_price_per_dish),
    expected_lb_meat_portion = as.numeric(expected_lb_meat_portion),
    oz_meat_per_dish = as.numeric(oz_meat_per_dish),
    portion_size_oz = as.numeric(portion_size_oz),
    planned_weight_lbs = as.numeric(planned_weight_lbs),
    cost_recipe_per_portion = as.numeric(cost_recipe_per_portion)
  )

ingredient_prices <- ingredient_prices %>%
  mutate(
    ingredient = trimws(ingredient),
    category = trimws(category),
    conventional_price_lb = as.numeric(conventional_price_lb),
    sustainable_price_lb = as.numeric(sustainable_price_lb),
    default_sus = trimws(default_sus)
  )

ghg_equivalents <- ghg_equivalents %>%
  mutate(
    meat_type = trimws(meat_type),
    c_footprint_kg_c_per_kg_food = as.numeric(c_footprint_kg_c_per_kg_food)
  )

```

## Filtering to just one dining hall

If desired, filter to just one dining hall:

```{r}
meals <- meals %>%
  filter(dining_hall == "C3")

```


## Building optimization df

This will give you the df optimization_data which is ready to use

```{r}
# =========================================================
# Build optimization_data
# =========================================================

ingredient_summary <- meals %>%
  filter(!is.na(ingredient), ingredient != "") %>%
  group_by(ingredient) %>%
  summarise(
    baseline_freq = n(),
    percent_dish_meat = safe_mean(percent_meat),
    meat_price_per_dish = safe_mean(meat_price_per_dish),
    expected_lb_meat_portion = safe_mean(expected_lb_meat_portion),
    oz_meat_per_dish = safe_mean(oz_meat_per_dish),
    portion_size_oz = safe_mean(portion_size_oz),
    expected_portions = safe_mean(planned_portions),
    planned_weight_lbs = safe_mean(planned_weight_lbs),
    cost_recipe_per_portion = safe_mean(cost_recipe_per_portion),
    category_meals = first_nonmissing(category),
    .groups = "drop"
  )

ingredient_lookup <- ingredient_prices %>%
  distinct(ingredient, .keep_all = TRUE) %>%
  transmute(
    ingredient,
    category_prices = category,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus
  )

ghg_lookup <- ghg_equivalents %>%
  distinct(meat_type, .keep_all = TRUE) %>%
  transmute(
    category = meat_type,
    conventional_ghg_per_lb = c_footprint_kg_c_per_kg_food
  )

optimization_data <- ingredient_summary %>%
  left_join(ingredient_lookup, by = "ingredient") %>%
  mutate(
    category = coalesce(category_prices, category_meals)
  ) %>%
  left_join(ghg_lookup, by = "category") %>%
  select(
    ingredient,
    category,
    percent_dish_meat,
    meat_price_per_dish,
    expected_lb_meat_portion,
    oz_meat_per_dish,
    conventional_price_lb,
    sustainable_price_lb,
    default_sus,
    portion_size_oz,
    expected_portions,
    planned_weight_lbs,
    cost_recipe_per_portion,
    conventional_ghg_per_lb,
    baseline_freq
  )

optimization_data <- optimization_data %>%
  mutate(
    baseline_freq = as.numeric(baseline_freq)
  )

# -------------------------
# Quick diagnostics
# -------------------------
optimization_data %>%
  filter(
    is.na(expected_lb_meat_portion) |
    is.na(conventional_price_lb) |
    is.na(sustainable_price_lb) |
    is.na(default_sus) |
    is.na(conventional_ghg_per_lb)
  )

```
